---
title: "Data Exploration and Cleaning"
author: "Kaiju Morquecho"
output: html_document
---

# Load Libraries
```{r message=FALSE}
library(tidyverse)
library(here)
library(janitor) 
library(lubridate)
```


# Read in data 
```{r message=FALSE}
asthma_emergencies <- read_csv(here("data","annual_asthma_emergencies_county_2020_2021_2022.csv")) %>%
  clean_names() %>%
  select(-data_comment, -x8) # drop unnecessary cols

annual_county_aqi_2020 <- read_csv(here("data","annual_aqi_by_county_2020.csv")) %>%
  clean_names() 

annual_county_aqi_2021 <- read_csv(here("data","annual_aqi_by_county_2021.csv")) %>%
  clean_names()

annual_county_aqi_2022 <- read_csv(here("data","annual_aqi_by_county_2022.csv")) %>%
  clean_names()

```

# Bind 2020, 2021 and 2022 AQI datasets 
```{r}
# bind all 3 years into one df
annual_aqi <- bind_rows(
  annual_county_aqi_2020,
  annual_county_aqi_2021,
  annual_county_aqi_2022
)
```

# Take a quick look at datasets
```{r}
glimpse(annual_aqi)
glimpse(asthma_emergencies)


# check for immediately obvious NAs or blanks
asthma_emergencies %>% filter(is.na(state) | state == "")
asthma_emergencies %>% filter(is.na(county) | county == "")

annual_aqi %>% filter(is.na(county) | county == "")
annual_aqi %>% filter(is.na(state) | state == "")

```

# Check geographic extent of each dataset
```{r}
# state
asthma_emergencies %>%
  select(state) %>%
  unique() %>%
  nrow() # 27 states

annual_aqi %>%
  select(state) %>%
  unique() %>%
  nrow() # all states and 4 territories

# identify states covered by both dfs
intersect(annual_aqi$state, asthma_emergencies$state)  # 25


# county 
asthma_emergencies %>%
  select(county) %>%
  unique() %>%
  nrow() # 995 counties covered

annual_aqi %>%
  select(county) %>%
  unique() %>%
  nrow() # 789 counties covered

# identify states covered by both dfs 
intersect(annual_aqi$county, asthma_emergencies$county) # 515
```
# Clean datasets

### 'value' column in asthma_emergencies
```{r}
# check class of 'value' col
class(asthma_emergencies$value) # need to change class from char to numeric 

# fix issues with 'value' column first - remove "," and "suppressed"
sum(str_detect(asthma_emergencies$value, ",")) # 265 rows written as 123,45 --> should be 12345

asthma_emergencies <- asthma_emergencies %>%
  mutate(value = str_remove(value, ","))

# recheck for commas
sum(str_detect(asthma_emergencies$value, ",")) # now 0

# check how many rows with 'suppressed'
sum(asthma_emergencies$value == 'Suppressed') # 260 

# filter out 'Suppressed' since no value can be recovered and row otherwise not useful
asthma_emergencies <- asthma_emergencies %>%
  filter(!value == "Suppressed")

# recheck 'Suppressed' is gone
sum(asthma_emergencies$value == 'Suppressed') # now 0 

# rename col with # of ashtma ER visits from 'value' to something meaningful and coerce from class character to numeric
asthma_emergencies <- asthma_emergencies %>%
  mutate(value = as.numeric(value)) %>%
  rename(er_visits = value)

# recheck class 
class(asthma_emergencies$er_visits) # numeric
```

### 'county_fips' and 'state_fips' columns in asthma_emergencies
```{r}
# county_fips
class(asthma_emergencies$county_fips) # char, leave as is to protect leading 0s

# check if any NAs
sum(is.na(asthma_emergencies$county_fips)) # 0


# state_fips
class(asthma_emergencies$state_fips) # char, leave as is to protect leading 0s

# check if any NAs
sum(is.na(asthma_emergencies$state_fips)) # 0
```
### 'state' and 'county' columns (in both)
```{r}
# trim off whitespace, if any, for county and state
asthma_emergencies <- asthma_emergencies %>%
  mutate(state = str_trim(str_to_title(state)),
         county = str_trim(str_to_title(county)))

annual_aqi <- annual_aqi %>%
  mutate(state = str_trim(str_to_title(state)),
         county = str_trim(str_to_title(county)))
```

### 'year' column (in both)
```{r}
# class 
class(asthma_emergencies$year) # numeric, leaving as is for robustness

class(annual_aqi$year) # numeric, leaving as is for robustness

```
### Check for total of NAs (in both)
```{r}
sum(is.na(asthma_emergencies)) # 0 NAs

sum(is.na(annual_aqi)) # 0 NAs

```

### Check all 15 numeric cols for NAs, negatives, and max # days
```{r}
# Check for NAs in numeric columns
annual_aqi %>%
  summarise(across(where(is.numeric), ~ sum(is.na(.))))  # 0 NA 

# Check for negative values in numeric columns
annual_aqi %>%
  summarise(across(where(is.numeric), ~ sum(. < 0, na.rm = TRUE)))  # no negative values

# check max values in cols where max and min should be 0 - 366 because unit is days out of a year
annual_aqi %>%
  summarise(across((contains('days')), ~ max(., na.rm = TRUE))) %>%
  count()

# drop row > 366 
annual_aqi <- annual_aqi %>%
  filter(if_all(contains("day"), ~ . <= 366))

```



### 'max_aqi' col in annual_aqi
```{r}
summary(annual_aqi)

# AQI is categorized up to 500, 500 > is beyond AQI scale but not impossible  
# all rows with max_aqi > 500
high_aqi <- annual_aqi %>% filter(max_aqi > 500) 

# states with max_aqi between 500 and 3000
high_aqi %>%
  filter(max_aqi <= 3000) %>% 
  
  # of those above 500, all but 1 row are below 3000. 
  # states with AQIs between 500 and 3000 experience       
  # severe wildfires every year
  
  count(state, sort = TRUE)


# drop the one outlier, row that is > 3000. # one row = 3000, an AQI this high (8368) is likely an error 

annual_aqi <- annual_aqi %>%
  filter(max_aqi <= 3000 )

# recheck row was dropped
annual_aqi %>% filter(max_aqi > 3000)

```












